Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-26 manuscript/paper.tex
revision.. TODO: results

@@

We consider
rigid rotation and tilting of the 
%BO$_{6/2}$ 
octahedra
 in ABO$_3$ perovskites
 from the viewpoint of Jahn-Teller problem
 and study the
effect of the crystal field of the nearest A and B ion cages 
%that immediately surround the octahedra
on the single electron spectrum of B, the 
central 
transition metal ions.
While the crystal field of the octahedra 
that 
% quenches the orbital angular momentum of the central B ion and 
creates degenerate manifolds $e_g$ and $t_{2g}$
do not change with such rigid rotations, 
the field of 
%A and B ionic cages %that immediately surround the octahedra
A and B cages
deforms significantly %with the octahedral rotations 
from the ideal octahedral field
in a way that
leads to sizeable splittings in 
%the $e_g$ and $t_{2g}$ manifolds of the central B ion.
these manifolds.
%Similar to the Jahn-Teller problem,
The lowering of the ground state energy
of a given many-electron configuration
due to these splittings
can thus be an important driving force behind the octahedral rotations 
and associated structural transformations.
We find that the size of the splitting and the 
%local orbital ordering 
order of the orbitals
are
determined by a competition between the A and B cages, 
as well as a competition between the tilt and rotation angles.

@@

%Their crystal structures 
The perovskite structures
have a common formula unit ABX$_3$,
where 
%positive ions A and negative ions X form a cube and an octahedron around positive ion B.
A and B are positive ions while X are negative ions 
%whose six-fold coordination forms octahedra around B.
that form octahedra around  B.
In a cubic perovskite,
the octahedra are symmetric, 
but they can 
distort~\cite{
BersukerCCR75,
KugelSPU82,
MillisN98,
TerakuraPMS07,
HalcrowRSC13,
LufasoAC04}
or simply 
rotate~\cite{
GlazerAC72, ThomasAC96,
WoodwardAC97a,WoodwardAC97b,
HowardAC98,StokesAC02,AngelPRL05}
 to lower the symmetry
and produce, e.g., a rhombohedral, tetragonal or orthorhombic structure.
% cubic perovskite [spacegroup 221] can transform to:
% crystal system: spacegroups
% triclinic: 2
% monoclinic: 11,12,15
% orthorhombic: 59,62,63,71,74
% tetragonal 127,139,140
% rhombohedral/trigonal: 167
% cubic: 204, 221 
Octahedral rotations also occur in other perovskite related structures
and
have been extensively 
studied
due to their crucial role in determining various physical properties,
e.g., in 
ferroelectric~\cite{GhosezNM11,
RondinelliAM12,
SchickPRB14,
SimPRB14,
LeePRL17,
WangPRL22,
ZhouPRB22,
GaoPRB23} and 
multiferroic~\cite{KhomskiiPhys09,
BenedekPRL11,
YangPRL14,
YePRB18,
ZhangPRL20,
ZhangYajunPRL20, 
ZhouPRB21,
ZhangPRL22}
systems.
The interplay of octahedral rotations
and 
spin-orbit coupling~\cite{AmatNL14,KrachJPCM23},
breathing distortions~\cite{BalachandranPRB13},
strain~\cite{Aguado-PuentePRL11},
defects~\cite{JiaPRB22}, and
pressure~\cite{XiangPRB17},
and 
their role in
magnetostructural phase transitions~\cite{JohnsonPRL20},
orbital order~\cite{OgawaPRL12,ChenPRL19}, orbital anisotropy~\cite{HuangPRB21}, 
and
%surface electronic structure under in-plane uniaxial compression~\cite{MoralesPRL23}
surface electronic structure~\cite{MoralesPRL23},
has also been investigated.
% JT interacting with octa tilting:
Since, usually, the octahedra are not just %(Jahn-Teller) 
distorted or rotated but a combination of both,
the influence of these two 
structural mutations %instabilities 
on each other has recently been explored by many authors~\cite{LufasoAC04,CarpenterAC09,ChenPRL19}
as well.

@@

Based on density functional theory (DFT) calculations,
a few recent works
~\cite{garcia-fernandezJPCL10,CammarataJCP14,LeeCM16,YoshidaPRL21}, 
however, 
suggest that the octahedral rotation and tilting is 
a second order Jahn-Teller effect
~\cite{HerzbergZPC33,BersukerPL66,KristoffelPSSB67,PearsonJMS83,FultonJCP04,BersukerCR13}
%\az{describe SOJT in a sentence: coupling to the low lying unoccupied states on the JT ion, or orbital hybridisation with other ions, coupling between bands...}
(i.e., the distortion is stabilised by a coupling between the ground and excited adiabatic potential energy surfaces).
% garcia-fernandezJPCL2010
Garcia-Fernandez~et~al.
study
potassium fluoride perovskite family KMF$_3$ (M$=$Ca$^{2+}$ or a $3$d transition metal)
and found that the tilting angle
depends on the covalent bonding through the mixing of occupied and 
low-lying unoccupied orbitals of the transition metal~\cite{garcia-fernandezJPCL10}.
% CammarataJCP2014
Cammarata and Rondinelli
find a negative correlation between metal-oxygen bond covalency 
and octahedral rotation in orthorhombic perovskites oxides~\cite{CammarataJCP14}.
% Lee, nick2016
Lee~et~al. find 
similar results for metal-iodine bonding in
lead iodide perovskites,
where the role of 
the hydrogen bonding 
in a hybrid perovskite with
organic ammonium ions at A sites
is also found to be crucial~\cite{LeeCM16}.
Yoshida~et~al.
use DFT along with 
representation theory analysis 
and 
indicate that 
the rotations are induced by bonding interaction 
between the valence and conduction band Bloch states at different wavevectors
that are mapped onto each other on Brillouin zone folding
 for the associated lattice distortion~\cite{YoshidaPRL21}.

@@

Beyond the anion octahedron,
each transition metal ion B in a cubic perovskite
is coordinated by two cation cages,
a cube of A and an octahedron of B, as shown in Figs.~\ref{fig:struc}(c,d).
Including the 
%electrostatic 
potential of these two ion cages
into the 
crystal field Hamiltonian of 
the central ion B,
we explore the effect 
of the rotation and tilting on 
%the splittings in 
its $t_{2g}$ and $e_g$ manifolds
that are
created primarily by the anion's octahedral potential.
We find that the degeneracy of $t_{2g}$ and $e_g$
is lifted, and
the size of the splitting 
and 
%local orbital ordering
the order of the orbitals
depend
on the relative charges on the two types of cations
and a competition between rotation and tilting.
At typical parameter values,
the splitting is large enough 
to play a major role in 
the corresponding structural instability. 
The qualitative picture developed by our calculations 
holds even when the potential of the full lattice is considered.

@@

The paper is organised as follows.
Section~\ref{sec:formalistm}
presents the formalism and 
sec.~\ref{sec:JT} reviews the Jahn-Teller effect to
introduce the notation and concepts involved
by considering tetragonal distortion of an octahedron.
%\az{[.... multiple subsections?]}
The model and calculations are presented
in sec.~\ref{sec:calc},
where
we first discuss the relative contributions of the three closest ion cages
in the cubic structure
 to
the crystal field at the central transition metal site 
and crystal field splitting between its $e_g$ and $t_{2g}$ manifolds
(sec.~\ref{sec:relativeVoctABO}),
and then 
describe the deformation effects of 
rigid octahedral rotations (sec.~\ref{sec:rotation})
on
the structure (sec.~\ref{sec:structure}) 
and crystal field (sec.~\ref{sec:deformCF}).
The results are in sec.~\ref{sec:results},
where we %seperately
present the splitting in the single particle spectrum
belonging to the $t_{2g}$ (sec.~\ref{sec:t2g})
and $e_g$ (sec.~\ref{sec:eg-gen}) manifolds,
and finally describe the effect of the full lattice (sec.~\ref{sec:FL}).
%\az{appendix?}

@@

If we have an arrangement of ions with charge $q$ around a given site,
we can express their (crystal field) potential in terms of spherical harmonics~\cite{PavariniChap}, 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{l,m}~\left(\frac{r}{a}\right)^l \ylm{l,m},
\end{align}
where $a$ sets the length scale
and $V_{l,m}$ are multipole components of the potential
 (for the given site as origin and a chosen orientation).
%and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we have a transition metal atom/ion
at the central site
with
only $d$-orbitals ($l=2$),
the matrix elements of its crystal field Hamiltonian
are then given by,
\begin{align}
H_{m',m''} &= -\frac{q}{a}\sum_{l,m} V_{l,m} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2,m'}| Y_{l,m}| Y_{2,m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$ is the expectation value for the $d$-orbitals, assumed to be the same for all $d$-orbitals, i.e., the radial part of all $d$-orbitals is assumed the same~\cite{PaxtonNotes}.
The sum is restricted to even values of $l\leq 4$ as 
$C_{l}^{m'm''m}=0$ otherwise.

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six oxygen (O) ions each with a charge $-q_o$ making an octahedron of size $a$ (placed at the face centres of a cube of edge length $a$).
The non-spherical part of the potential they
 produce at the centre of the octahedron
 %that is 
 relevant for the $d$-orbitals contains 
 only two $l=4$ components
  in a specific ratio (${V_{4,0}:V_{4,4}=1:\sqrt{5/7}}$),
 and is called octahedral field. 
It is given by~\cite{PavariniChap},
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B 
%at the centre 
in the field $V_{oct}(\vec r)$
turns out to be diagonal,
with its states split into two manifolds: 
% $\Delta$ apart:
$H_{m,m} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{m,m} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

@@

The dependence of the elastic energy on the distortion can be roughly explained
as follows.
Consider an 
elongation or a compression of an octahedron along one of its axes by $a\delta$,
where $a$ is the lattice constant of the cubic structure.
The corresponding change in the unit cell volume would be linear in $\delta$
as $\sim a^3 \delta$
so the elastic energy
would change as $\sim\delta^2$.
Assuming a linear change 
 in the electronic spectrum,
  the total energy has a form
$E=-\rho\delta + K \delta^2$, where $\rho$ and $K$ are positive constants 
describing the changes in the electronic and elastic energies,
and would be optimum at $\delta=\rho/2K$.
In general, as long as the two competing terms 
are not the same degree monomial in the distortion,
the ground state can have a finite distortion.
We use this argument to focus only on the changes
 in the electronic energy throughout this paper.

@@

\subsection{Tetragonal distortion of an octahedron}

Let us consider elongation of an octahedron in sec.~\ref{sec:egt2g} 
along one of its O-B-O bond. 
Align the cartesian axes along O-B-O bonds and stretch the two bonds along $z$-axis equally by an amount $a\delta$.
%where $a$ is the size of the octahedron (the distance between the O-O along any cartesian axes).
The potential $V^{JT}(\vec r)$ at the central B site (JT ion) of this deformed octahedron 
still contains only a few non-zero components.
Not only the ratio 
$V_{4,0}^{JT}:V_{4,4}^{JT}$ now deviates from
 ideal 
octahedral field as in Eq.~\ref{eq:voct},
there is an $l=2$ component induced due to the deformation,
\begin{align}
%V_{0,0}^{JT}=& \frac{8 \sqrt{\pi } (2 \delta +3)}{\delta +1},\\
V_{2,0}^{JT}=& -\frac{32 \sqrt{\frac{\pi }{5}} \delta  \left(\delta ^2+3 \delta +3\right)}{(\delta +1)^3},\\
V_{4,0}^{JT}=& \frac{32}{3} \sqrt{\pi } \left(\frac{4}{(\delta +1)^5}+3\right),\\
V_{4,4}^{JT}=& \frac{32 \sqrt{35 \pi }}{3}.
\end{align}

@@

The crystal field Hamiltonian of the central B ion/JT ion
${H_{m',m''}^{JT} = (q_o/a)\sum_{l,m} V_{l,m}^{JT} C_{l}^{m'm''m} D_{l}}$
still turns out to be diagonal 
(i.e., the deformation $\delta$
will not couple these orbitals to each other),
but the degeneracies of the two manifolds is lifted
and their respective average energies also shifted.
Defining 
$F_n(\delta) = 1-1/(1+\delta)^n$,
we see that
the $t_{2g}$ now splits into two levels,
with energies
\begin{align}
E_{xy} =&
\frac{32}{7}\frac{q_o}{a} \left(
 D_2F_3 - \frac{20}{9}D_4F_5 
\right),\\
E_{yz} =& E_{zx} = -\frac{1}{2} E_{xy},
\end{align}
as measured from their average 
$E_0^{t_{2g}} = -64q_oD_4 \left(3-F_5\right)/9a$.
Similarly, $e_g$ splits up as
\begin{align}
E_{x^2-y^2}=&
\frac{32}{7}\frac{q_o}{a} \left(D_2F_3+\frac{5}{3} D_4F_5\right),\\
E_{z^2}=&-E_{x^2-y^2},\\
\end{align}
as measured from their average 
$E_0^{e_{g}} = 32 q_oD_4 \left(3-F_5\right)/3a$.
The crystal field splitting becomes
$\Delta \equiv E_0^{e_{g}}-E_0^{t_{2g}}=
160q_oD_4\left(3-F_5\right)/9a$.
At small $\delta$, 
expanding 
$F_3,F_5$
 up to first order in $\delta$,
we obtain
\begin{align}
\label{eq:JTt2g}
(E_{xy},E_{yz},E_{zx})&= (2, -1, -1)
%\left[\frac{16}{63} (27 \text{D2}-100 \text{D4})\right]
\left[D_2-\frac{100}{27}  D_4\right]\frac{48}{7}\frac{q_o}{a}\delta,\\
\label{eq:JTeg}
(E_{x^2-y^2},E_{z^2})&=(1,-1)
\left[D_2+\frac{25}{9} D_4\right]\frac{96}{7}\frac{q_o}{a}\delta,
\end{align}
whereas
%${\Delta_{\delta}= \frac{48}{7}\frac{q_o}{a}\frac{\delta}{a}}$.
$E_0^{t_{2g}}$, $E_0^{e_{g}}$ and $\Delta$ all are rescaled by a factor of ${1-5\delta/3}$.

@@

We now turn to our model and calculations of the splittings in the 
$t_{2g}$ and $e_{g}$ manifolds due to the deformations produced 
in the crystal field as the octahedra rotate or tilt.
First, let us consider the cubic perovskite and see 
how much A and B cages contribute to the octahedral crystal field and associated splitting $\Delta$ between $t_{2g}$ and $e_{g}$ manifolds.

@@

Including the contribution of these AB cages to the crystal field,
we obtain
\begin{align}
%V(\vec r) &= V_{O}(\vec r) + V_{AB}(\vec r),\\
V(\vec r)  &= (1+f_{q_A}) V_{O}(\vec r),
\end{align}
whereas the 
 splitting becomes
%\begin{align}
$\Delta = (1+f_{q_A})\Delta_o$.
%\end{align}
%where ${\Delta_O=\frac{160}{3}\frac{q_o}{a} D_{4}}$.
We see that, relative to the O octahedron,
 the contribution of the
 two cation cages towards the total field splitting
 is relatively small at any $q_A$ but
 still large enough to be ignored.
It can be measured by $(\Delta-\Delta_o)/\Delta_o=f_{q_A}$,
which linearly changes from 
$-\frac{3}{32}\simeq -10\%$ at $q_A=0$
to $\frac{8}{27 \sqrt{3}}\simeq 17\%$ at $q_A=3q_o$. 
Typical values of $q_A$ relevant for real transition metal oxides
can be safely assumed to range between $0$ and $4$;  
$q_A=0$ for systems like tungsten oxide $WO_3$ that do not have any cations at the A sites,
and $q_A=4$ still contributing just above $8\%$.
We will see that these with the octahedral rotations and tilting, 
these potentials rotate and deform.
The deformation also induces $l=2$ components, which couple to the 
$d$-orbitals of the B ion much more strongly. 
%due to a much larger $D_{l=2}$.

\subsection{Rigid octahedral rotations}
\label{sec:rotation}

\subsection{Transformation of the crystal structure}
\label{sec:structure}

@@

Since $V_{O}(\vec r)$ is relatively stronger, 
we can take the octahedra's local frames (that are rotated with them)
to first describe the effect of $V_{O}(\vec r)$, i.e.,
splitting of degenerate d-orbitals into $e_g, t_{2g}$ manifolds with 
a gap $\Delta_o$,
and then
consider the effect of $V_{AB}(\vec r)$ on these manifolds.
$V_{AB}(\vec r)$ will couple and mix the orbitals in the two manifolds.
Since %$\Delta_AB << \Delta_O$,
$f_{q_A} \ll 1$,
we can even ignore the coupling between the states belonging to the 
different manifolds $\Delta_o$ apart.
This also makes it possible to analytically solve the problem in two limiting cases discussed later.

>>>>

\author{Kashif Sabeeh}
\affiliation{Department of Physics, Quaid-i-Azam University, Islamabad 45320, Pakistan}

@@

We consider
rigid rotation and tilting of the 
%BO$_{6/2}$ 
anion octahedra around the transition metal ions
 in ABO$_3$ perovskites
 from the viewpoint of Jahn-Teller problem
 and study the
effect of the crystal field of the nearest A and B ion cages 
on the single electron spectrum of the 
central 
transition metal B ions.
While the crystal field of the octahedra,
which
% quenches the orbital angular momentum of the central B ion and 
creates degenerate manifolds $e_g$ and $t_{2g}$,
do not change with such rigid rotations, 
the field of 
%A and B ionic cages %that immediately surround the octahedra
A and B cages
deforms significantly %with the octahedral rotations 
from the ideal octahedral field
in a way that
leads to sizeable splittings in 
%the $e_g$ and $t_{2g}$ manifolds of the central B ion.
these manifolds.
%Similar to the Jahn-Teller problem,
The lowering of the ground state energy
of a given many-electron configuration
due to these splittings
can thus be an important driving force behind the octahedral rotations 
and associated structural transformations.
We find that the size of the splitting and the 
%local orbital ordering 
order of the orbitals
are
determined by a competition between the A and B cages, 
as well as a competition between the tilt and rotation angles.

@@

%Their crystal structures 
The perovskite structures
have a common formula unit ABX$_3$,
where 
%positive ions A and negative ions X form a cube and an octahedron around positive ion B.
A and B are positive ions while X are negative ions 
%whose six-fold coordination forms octahedra around B.
that form octahedra around  B.
In a cubic perovskite,
the octahedra are symmetric, 
but they can 
distort~\cite{
BersukerCCR75,
KugelSPU82,
MillisN98,
TerakuraPMS07,
HalcrowRSC13,
LufasoAC04}
or simply 
rotate~\cite{
GlazerAC72, ThomasAC96,
WoodwardAC97a,WoodwardAC97b,
HowardAC98,StokesAC02,AngelPRL05}
 to lower the symmetry
and produce, e.g., a rhombohedral, tetragonal or orthorhombic structure.
% cubic perovskite [spacegroup 221] can transform to:
% crystal system: spacegroups
% triclinic: 2
% monoclinic: 11,12,15
% orthorhombic: 59,62,63,71,74
% tetragonal 127,139,140
% rhombohedral/trigonal: 167
% cubic: 204, 221 
Octahedral rotations also occur in other perovskite related structures
and
have been extensively 
studied
due to their crucial role in determining various physical properties,
e.g., in 
ferroelectric~\cite{GhosezNM11,
RondinelliAM12,
SchickPRB14,
SimPRB14,
LeePRL17,
WangPRL22,
ZhouPRB22,
GaoPRB23} and 
multiferroic~\cite{KhomskiiPhys09,
BenedekPRL11,
YangPRL14,
YePRB18,
ZhangPRL20,
ZhangYajunPRL20, 
ZhouPRB21,
ZhangPRL22}
systems.
The interplay of octahedral rotations
and 
spin-orbit coupling~\cite{AmatNL14,KrachJPCM23},
breathing distortions~\cite{BalachandranPRB13},
strain~\cite{Aguado-PuentePRL11},
defects~\cite{JiaPRB22}, and
pressure~\cite{XiangPRB17},
and 
their role in
magnetostructural phase transitions~\cite{JohnsonPRL20},
orbital order~\cite{OgawaPRL12,ChenPRL19}, orbital anisotropy~\cite{HuangPRB21}, 
and
%surface electronic structure under in-plane uniaxial compression~\cite{MoralesPRL23}
surface electronic structure~\cite{MoralesPRL23},
have also been investigated.
% JT interacting with octa tilting:
Since, usually, the octahedra are not just %(Jahn-Teller) 
distorted or rotated but a combination of both,
the influence of these two 
structural mutations %instabilities 
on each other has recently been explored by many authors~\cite{LufasoAC04,CarpenterAC09,ChenPRL19}
as well.

@@

Based on density functional theory (DFT) calculations,
a few recent works
~\cite{garcia-fernandezJPCL10,CammarataJCP14,LeeCM16,YoshidaPRL21}, 
however, 
suggest that the octahedral rotation and tilting is 
a second order Jahn-Teller effect
~\cite{HerzbergZPC33,BersukerPL66,KristoffelPSSB67,PearsonJMS83,FultonJCP04,BersukerCR13}
%\az{describe SOJT in a sentence: coupling to the low lying unoccupied states on the JT ion, or orbital hybridisation with other ions, coupling between bands...}
(i.e., the distortion is stabilised by a coupling between the ground and excited adiabatic potential energy surfaces).
% garcia-fernandezJPCL2010
Garcia-Fernandez~et~al.
study
potassium fluoride perovskite family KMF$_3$ (M$=$Ca$^{2+}$ or a $3$d transition metal)
and found that the tilting angle
depends on the covalent bonding through the mixing of occupied and 
low-lying unoccupied orbitals of the transition metal~\cite{garcia-fernandezJPCL10}.
% CammarataJCP2014
Cammarata and Rondinelli
find a negative correlation between metal-oxygen bond covalency 
and octahedral rotation in orthorhombic perovskite oxides~\cite{CammarataJCP14}.
% Lee, nick2016
Lee~et~al. find 
similar results for metal-iodine bonding in
lead iodide perovskites,
where the role of 
the hydrogen bonding 
in a hybrid perovskite with
organic ammonium ions at A sites
is also found to be crucial~\cite{LeeCM16}.
Yoshida~et~al.
use DFT along with 
representation theory analysis 
and 
indicate that 
the rotations are induced by bonding interaction 
between the valence and conduction band Bloch states at different wavevectors
that are mapped onto each other on Brillouin zone folding
 for the associated lattice distortion~\cite{YoshidaPRL21}.

@@

Beyond the anion octahedron,
each transition metal ion B in a cubic perovskite
is coordinated by two cation cages,
a cube of A and an octahedron of B, as shown in Figs.~\ref{fig:struc}(c,d).
Including the 
%electrostatic 
potential of these two ion cages
into the 
crystal field Hamiltonian of 
the central ion B,
we explore the effect 
of the rotation and tilting on 
%the splittings in 
its $t_{2g}$ and $e_g$ manifolds
that are
created primarily by the anion's octahedral potential.
We find that the degeneracy of $t_{2g}$ and $e_g$
is lifted. 
The size of the splitting 
and 
%local orbital ordering
the order of the orbitals
depend
on the relative charges on the two types of cations
and a competition between rotation and tilting.
At typical parameter values,
the splitting is large enough 
to play a major role in 
the corresponding structural instability. 
The qualitative picture developed by our calculations 
holds even when the potential of the full lattice is considered.

@@

The paper is organised as follows.
Section~\ref{sec:formalistm}
presents the formalism and 
sec.~\ref{sec:JT} reviews the Jahn-Teller effect to
introduce the notation and concepts involved
by considering tetragonal distortion of an octahedron.
%\az{[.... multiple subsections?]}
The model and calculations are presented
in sec.~\ref{sec:calc},
where
we first discuss the relative contributions of the three closest ion cages
in the cubic structure
 to
the crystal field at the central transition metal site 
and crystal field splitting between its $e_g$ and $t_{2g}$ manifolds
(sec.~\ref{sec:relativeVoctABO}),
and then 
describe the deformation effects of 
rigid octahedral rotations %(sec.~\ref{sec:rotation})
on
the structure (sec.~\ref{sec:structure}) 
and crystal field (sec.~\ref{sec:deformCF}).
The results are in sec.~\ref{sec:results},
where we %seperately
present the splitting in the single particle spectrum
belonging to the $t_{2g}$ (sec.~\ref{sec:t2g})
and $e_g$ (sec.~\ref{sec:eg-gen}) manifolds,
and finally describe the effect of the full lattice (sec.~\ref{sec:FL}).
%\az{appendix?}

@@

If we have an arrangement of ions with charge $q$ around a given site,
we can express their %(crystal field) 
potential in terms of spherical harmonics~\cite{PavariniChap}, 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{l,m}~\left(\frac{r}{a}\right)^l \ylm{l,m},
\end{align}
where $a$ sets the length scale
and $V_{l,m}$ are multipole components of the potential
 (for the given site as origin and a chosen orientation).
%and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we have a transition metal atom/ion
at the central site
with
only $d$-orbitals ($l=2$),
the matrix elements of its crystal field Hamiltonian
are then given by,
\begin{align}
H_{m',m''} &= -\frac{q}{a}\sum_{l,m} V_{l,m} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2,m'}| Y_{l,m}| Y_{2,m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$
is the expectation of $(r/a)^l$ evaluated in the $d$-orbitals bases
assuming the same radial dependence for all orbitals~\cite{PaxtonNotes}.
The sum is restricted to even values of $l\leq 4$ as 
$C_{l}^{m'm''m}=0$ otherwise,
with $-l \leq m \leq l$.

We consider transition metal oxides as an example perovskite system throughout the rest of the manuscript,
but our calculations and results are general and also apply to perovskite transition metal compounds with other anions, e.g., halides and chalcogenides.

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six oxygen (O) ions each with a charge $-q_o$ making an octahedron of size $a$ (where these anions are placed at the face centres of a cube of edge length $a$).
The non-spherical part of the potential they
 produce at the centre of the octahedron
 %that is 
 relevant for the $d$-orbitals contains 
 only two $l=4$ components, 
  in a specific ratio (${V_{4,0}:V_{4,4}=1:\sqrt{5/7}}$),
 and is called octahedral field. 
It is given by~\cite{PavariniChap},
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B 
%at the centre 
in the field $V_{oct}(\vec r)$
turns out to be diagonal,
with its states split into two manifolds: 
% $\Delta$ apart:
$H_{m,m} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{m,m} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

@@

The dependence of the elastic energy on the distortion can be roughly explained
as follows.
Consider an 
elongation or a compression of an octahedron along one of its axes by $a\delta$,
where $a$ is the size of the octahedron and lattice constant of the cubic perovskite structure.
The corresponding change in the unit cell volume would be linear in $\delta$
as $\sim a^3 \delta$
so the elastic energy
would change as $\sim\delta^2$.
Assuming a linear change 
 in the electronic spectrum,
  the total energy has a form
$E=-\rho\delta + K \delta^2$, where $\rho$ and $K$ are positive constants 
describing the changes in the electronic and elastic energies,
and would be optimum at $\delta=\rho/2K$.
In general, as long as the two competing terms 
are not the same degree monomial in the distortion,
the ground state can have a finite distortion.
We use this argument to focus only on the changes
 in the electronic energy throughout this paper.

@@

\subsection{Tetragonal distortion of an octahedron}
\label{sec:JTtertagonal}

Let us consider elongation of an octahedron % in sec.~\ref{sec:egt2g} 
along one of its O-B-O bonds. 
Align the cartesian axes along O-B-O bonds and stretch the two bonds along $z$-axis equally by an amount $a\delta$.
%where $a$ is the size of the octahedron (the distance between the O-O along any cartesian axes).
The potential $V^{JT}(\vec r)$ at the central B site (JT ion) of this deformed octahedron 
still contains only a few non-zero components.
Not only the ratio 
$V_{4,0}^{JT}:V_{4,4}^{JT}$ now deviates from
 ideal 
octahedral field in Eq.~\ref{eq:voct},
there is an $l=2$ component induced due to the deformation,
\begin{align}
%V_{0,0}^{JT}=& \frac{8 \sqrt{\pi } (2 \delta +3)}{\delta +1},\\
V_{2,0}^{JT}=& -\frac{32 \sqrt{\frac{\pi }{5}} \delta  \left(\delta ^2+3 \delta +3\right)}{(\delta +1)^3},\\
V_{4,0}^{JT}=& \frac{32}{3} \sqrt{\pi } \left(\frac{4}{(\delta +1)^5}+3\right),\\
V_{4,4}^{JT}=& \frac{32 \sqrt{35 \pi }}{3}.
\end{align}

@@

The crystal field Hamiltonian of the JT ion,
${H_{m',m''}^{JT} = (q_o/a)\sum_{l,m} V_{l,m}^{JT} C_{l}^{m'm''m} D_{l}}$,
still turns out to be diagonal 
(i.e., the deformation $\delta$
will not couple these orbitals to each other),
but the degeneracies of the two manifolds is lifted
and their respective average energies also shifted.
Defining 
$F_n(\delta) = 1-1/(1+\delta)^n$,
we see that
the $t_{2g}$ now splits into two levels,
with energies
\begin{align}
E_{xy} =&
\frac{32}{7}\frac{q_o}{a} \left(
 D_2F_3 - \frac{20}{9}D_4F_5 
\right),\\
E_{yz} =& E_{zx} = -\frac{1}{2} E_{xy},
\end{align}
as measured from their average 
$E_0^{t_{2g}} = -64q_oD_4 \left(3-F_5\right)/9a$.
Similarly, $e_g$ splits up as
\begin{align}
E_{x^2-y^2}=&
\frac{32}{7}\frac{q_o}{a} \left(D_2F_3+\frac{5}{3} D_4F_5\right),\\
E_{z^2}=&-E_{x^2-y^2},\\
\end{align}
as measured from their average 
$E_0^{e_{g}} = 32 q_oD_4 \left(3-F_5\right)/3a$.
The crystal field splitting becomes
$\Delta \equiv E_0^{e_{g}}-E_0^{t_{2g}}=
160q_oD_4\left(3-F_5\right)/9a$.
At small $\delta$, 
expanding 
$F_3,F_5$
 up to first order in $\delta$,
we obtain
\begin{align}
\label{eq:JTt2g}
(E_{xy},E_{yz},E_{zx})&= (2, -1, -1)
%\left[\frac{16}{63} (27 \text{D2}-100 \text{D4})\right]
\left[D_2-\frac{100}{27}  D_4\right]\frac{48}{7}\frac{q_o}{a}\delta,\\
\label{eq:JTeg}
(E_{x^2-y^2},E_{z^2})&=(1,-1)
\left[D_2+\frac{25}{9} D_4\right]\frac{96}{7}\frac{q_o}{a}\delta,
\end{align}
whereas
%${\Delta_{\delta}= \frac{48}{7}\frac{q_o}{a}\frac{\delta}{a}}$.
$E_0^{t_{2g}}$, $E_0^{e_{g}}$ and $\Delta$ all are rescaled by a factor of ${1-5\delta/3}$.

@@

We now turn to our model and calculations of the splittings in the 
$t_{2g}$ and $e_{g}$ manifolds due to the deformations 
in the crystal field that are produced as the octahedra rotate or tilt.
First, let us consider the cubic perovskite and see 
how much A and B cages contribute to the octahedral crystal field and associated splitting $\Delta$ between $t_{2g}$ and $e_{g}$ manifolds.

@@

Including the contribution of these AB cages to the crystal field,
we obtain
\begin{align}
%V(\vec r) &= V_{O}(\vec r) + V_{AB}(\vec r),\\
V(\vec r)  &= (1+f_{q_A}) V_{O}(\vec r),
\end{align}
whereas the 
 splitting becomes
%\begin{align}
$\Delta = (1+f_{q_A})\Delta_o$.
%\end{align}
%where ${\Delta_O=\frac{160}{3}\frac{q_o}{a} D_{4}}$.
We see that, relative to the O octahedron,
 the contribution of the
 two cation cages towards the total field splitting
 is relatively small at any $q_A$ but
 still large enough to be ignored.
It can be measured by $(\Delta-\Delta_o)/\Delta_o=f_{q_A}$,
which linearly changes from 
$-\frac{3}{32}\simeq -10\%$ at $q_A=0$
to $\frac{8}{27 \sqrt{3}}\simeq 17\%$ at $q_A=3q_o$. 
Typical values of $q_A$ relevant for real transition metal oxides
can be safely assumed to range between $0$ and $4$;  
$q_A=0$ for systems like tungsten oxide $WO_3$ that do not have any cations at the A sites,
and $q_A=4$ still contributing just above $8\%$.
We will see that,
as the anion octahedra rotate or tilt,
$V_{AB}(\vec r)$ rotate in its frame and deform as well.
The deformation also induces $l=2$ components, which couple to the 
$d$-orbitals of the B ion much more strongly. 
%due to a much larger $D_{l=2}$.

%\subsection{Rigid octahedral rotations}
%\label{sec:rotation}

\subsection{Rigid octahedral rotations: Transformation of the crystal structure}
\label{sec:structure}

@@

Since $V_{O}(\vec r)$ is relatively stronger, 
we can take the octahedra's local frames (that are rotated with them)
to first describe the effect of $V_{O}(\vec r)$, i.e.,
splitting of degenerate d-orbitals into $e_g, t_{2g}$ manifolds with 
a gap $\Delta_o$,
and then
consider the effect of $V_{AB}(\vec r)$ on these manifolds.
$V_{AB}(\vec r)$ will couple and mix the orbitals in the two manifolds.
Since %$\Delta_AB << \Delta_O$,
$f_{q_A} \ll 1$,
we can even ignore the coupling between the states belonging to
different manifolds that are $\Delta_o$ apart.
This also makes it possible to analytically solve the problem in two limiting cases discussed later.
